---
title: "Discrete Choice Models (DCM) for Assortative Mating (AM), Part 1: Simulating a Marriage Market"
description: "The first note in a small series: create actors, quasi-random locations, group attractors, and a utility-based residential assignment."
date: "2026-07-17"
author: "Benjamin F. Jarvis"
metadata-files:
  - ../_post-defaults.yml
categories:
  - "assortative mating"
  - "discrete choice models"
  - "log-linear models"
  - "simulation"
  - "spatial analysis"
bibliography: references.bib
---

## Introduction

There's a been a percolating trend in assortative mating research, particularly in demography and sociology, away from log-linear modeling and towards a more flexible style of discrete choice modeling, primarily based on Daniel McFadden's multinomial logit model.^[Confusingly, this is also referred to as the conditional logistic regression model. This model actually nests what sociologists and political scientists often call "multinomial logistic regression."] Examples include the work of Karen Haandrikman [-@haandrikmanExplainingFlightCupids2012], Aaron Gullickson [-@gullicksonCounterfactualChoiceApproach2021], and my colleague and student, Jesper Lindmarker [-@lindmarkerHoweverFarAway2025]. That the conditional logit model offers a more general formulation of the log-linear models that have served as the bread and butter of mobility and assortative mating research has been known for quite some time. John A. Logan showed as much in the 1980s [-@loganMultivariateModelMobility1983], as did Richard Breen in the 1990s [-@breenIndividualLevelModels1994]. Logan even points to knowledge of the equivalence dating back to the 1970s.

That the tiny revolution in assortative mating research away from traditional log-linear modeling to discrete choice modeling is taking place outside of economics, where discrete choice models were first developed, is perhaps not too surprising: sociologists are less constrained by issues of causal identification, general equilibrium, and stable matching assumptions that preoccupy economists. Using an out-of-the-box discrete choice modeling approach largely elides these issues in favor of arriving at parameter estimates that aren't so readibly interpreted in a causal or structural sense, but at least provide rich fodder for sociological interpretation.

So, this research note. My goal is to demonstrate, yet again, the equivalence between conditional logistic regression (i.e., McFadden's choice model) and log-linear analysis for relatively simple model specifications. But I want to go beyond that and also explore the greater possibilities, as well as the interpretive perils, that go along with this modeling approach. I do so using a simulated dataset representing some intuitive ideas about the factors affecting assortative mating.

This is just part one. This post sets the scene through simulation; stipulating and then allocating singles to social and spatial positions in a stylized marriage market. In subsequent posts, I'll show how to take these singles and produce matched pairs, how to fit log-linear and conditional logistic regression models to the resulting data, how the parameter estimates compare, and than how these parameter estimates might give us a skewed view of the "deeper" parameters that actually drive union formation in the simulated data.

## Building a simulated marriage market

I consider a marriage market characterized by a simplified geometric/geographic space, consisting of quasi-randomly generated positions that are occupied by actors with varying attributes who will then be paired. The pairing step will be covered in another post. This post is just about building the space, generating actors, and assigning actors to positions based on some simplified assumptions about the uneven distribution of different social groups.

### Positions

My goal is to create a semi-regular arrangement of what might be construed as housing units or residential positions for actors to occupy. These positions are static, and so the actors are conceived of as static. Their mobility is implicit in how space determines union formation, which I cover in the next post. This gives a geographic position to each actor that will influence subsequent matching. A simple way of doing this would be to use independent draws of a set of x and y coordinates from a uniform distribution. However, this creates a set of positions characterized by unusual gaps and overlaps that just doesn't look "right" to my eye. So instead of using pseudo-random numbers, I use quasi-random numbers, in this case Halton sequences, to generate the x and y coordinates. Quasi-random numbers tend to be more evenly spaced than pseudo-random numbers, which feels right for this fictive geography.

First things first a few libraries. I prefer `dplyr` for data handling and `ggplot` for plotting. I use `mnorm` for its halton sequence generation:

```{r}
library(dplyr)  
library(tidyr)
library(ggplot2)  
library(mnorm)
```

We can generate and compare the pseudo- and quasi-random positions as follows:

```{r}
#| label: halton-vs-uniform
#| fig-cap: "Quasi-geographic positions generated from Halton sequences and draws from a uniform distribution."
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.2
#| out-width: 100%

n<-800

xy_sequence <- halton(n = n, base = c(2, 3), start = 1L)

xy<-bind_rows(
tibble(
  position_id = seq_len(n),
  x = xy_sequence[, 1],
  y = xy_sequence[, 2],
  type="halton"
),
tibble(
  position_id = seq_len(n),
  x=runif(n),
  y=runif(n),
  type="uniform"
)
)

ggplot(xy,aes(x=x,y=y)) +
  geom_point(size = 0.75) +
  facet_wrap(vars(type),nrow = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = TRUE) +
  theme_void() + 
  theme(
    aspect.ratio = 1,
    plot.margin = margin(3, 3, 3, 3)
  )

```

As I said, the Halton draws just look a bit better for the present purposes. It resembles a sort of `social distancing` heuristic, with positions filling out the space in an irregular, but not overly clumpy way. I save the position generator in a function, and then call it to produce positions.

```{r}
make_positions <- function(n) {
  xy_sequence <- mnorm::halton(n = n, base = c(2, 3), start = 1L)
  tibble(
    position_id = seq_len(n),
    x = xy_sequence[, 1],
    y = xy_sequence[, 2]
  )
}
xy<-make_positions(n)
```

### Actors

For the actors, I assume they are split into two genders, men and women, with sebsequent unions constituted by opposite gendered pairs. I'll make the market gender balanced for the sake of simplicity, and when all is said and done, each actor will be paired with one and only one opposite gender partner. Beyond gender and (eventually) position, I stipulate that each actor is characterized by just two other attributes: group membership, proxying something like race or ethnicity, and age, with the possibility of different age distributions for each group.

I use scaled beta distributions to generate the ages. This allows me to constrain the ages to fall between two bounds. This is represented in the following function:

```{r}
scale_beta <- function(n, mean, precision, min_value = 0, max_value = 1) {
  scaled_mean <- (mean - min_value) / (max_value - min_value)
  shape1 <- scaled_mean * precision
  shape2 <- (1 - scaled_mean) * precision
  min_value + (max_value - min_value) * rbeta(n, shape1, shape2)
}
```

With this function in hand, I can create a dataset of simulated actors, for whom I also generate identifiers, which will be convenient in subsequent posts. The mean ages of the groups differ, but the age ranges are the same.

```{r}
scale_beta <- function(n, mean, precision, min_value = 0, max_value = 1) {
  scaled_mean <- (mean - min_value) / (max_value - min_value)
  shape1 <- scaled_mean * precision
  shape2 <- (1 - scaled_mean) * precision
  min_value + (max_value - min_value) * rbeta(n, shape1, shape2)
}

actors <- 
  tibble(
    group = c("a", "b", "c"),
    age_mean = c(25, 30, 35),
    n = c(200, 120, 80)
  ) |>
  expand_grid(tibble(sex = c("man", "woman"))) |> 
  reframe(
    person = seq_len(n),
    age = scale_beta(n, age_mean, precision = 7, min_value = 18, max_value = 45),
    .by = c(group, age_mean, sex)
  ) |>
  ungroup() |>
  mutate(id = row_number()) |> 
  select(id, sex, group, age)

slice_sample(actors,n=10)
```

The differing sampled age distributions, broken down separately for men and women, can be visualized using a density plot:

```{r}
#| label: age-plot
#| fig-cap: "Groups differ in mean age, but the simulated age distributions overlap."
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.6
#| out-width: 100%
group_colors <- c(a = "#c73535", b = "#555c60", c = "#287fc4")
ggplot(actors, aes(age, fill = group)) +
  geom_density(alpha = 0.45, color = NA) +
  facet_wrap(~ sex) +
  scale_fill_manual(values = group_colors) +
  labs(x = "Age", y = "Density", fill = "Group") +
  theme_minimal(base_size = 12)
```

### Assigning actors to locations

I want to introduce some group and age segregation into the simulated marriage market. I accomplish this by designating a set of attraction points. A point in the center pulls in younger actors. Then a set of three, group-specific poles outside the center pull in members of the respective group. 

```{r}
#| label: attractor-plot
#| fig-cap: "Group-specific attractors and the central youth attractor."
#| fig-align: center
#| fig-width: 6.6
#| fig-height: 6.6
#| out-width: 86%

poles <- tibble(
  group = c("a", "b", "c","youth"),
  group_x = c(0.25, 0.50, 0.75, 0.5),
  group_y = c(0.25, 0.75, 0.25, 0.5)
)

ggplot() +
  geom_point(
    data = xy,
    aes(x, y),
    size = 0.8,
    alpha = 0.28,
    color = "#9aa0a3"
  ) +
  geom_point(
    data = poles,
    aes(group_x, group_y, color = group),
    size = 5
  ) +
  geom_text(
    data = poles,
    aes(group_x, group_y, label = group, color=group),
    nudge_y = 0.055,
    size = 8,
    show.legend = FALSE
  ) +
  coord_equal(xlim = c(-0.03, 1.03), ylim = c(-0.03, 1.03), expand = FALSE) +
  scale_color_manual(values = group_colors) +
  labs(x = "Spatial coordinate x", y = "Spatial coordinate y", color = "Group") +
  theme_void(base_size = 12) + 
  theme(
    legend.position="none",
    plot.margin = margin(4, 4, 4, 4)
  )
```


To assign actors to location, I use utility-based probabilistic assignment approach. The utility function for each actor is based on the distance to the actor's respective group pole and on the distance to the youth attraction point, with the latter depending on the age of the actor him/herself. A small, random noise term is also added to each utility. Then iterate through locations, assigning 

```{r}
noise_scale<-2
utilities <- actors |>
  select(id, group, age) |>
  crossing(xy) |>
  left_join(poles, by = "group") |>
  mutate(
    youth_x = 0.5,
    youth_y = 0.5,
    youth_pull = (max(age) - age) / (max(age) - min(age)),
    group_distance = log(sqrt((x - group_x)^2 + (y - group_y)^2) + 0.01),
    youth_distance = log(sqrt((x - youth_x)^2 + (y - youth_y)^2) + 0.01),
    utility = -1 * group_distance - 2 * youth_pull * youth_distance
  ) |>
  ungroup() |>
  mutate(
    utility = utility + noise_scale * rnorm(n(), mean = 0, sd = sd(utility))
  ) |>
  arrange(desc(utility))

assigned_actors <- integer()
assigned_locations <- integer()
assignments <- vector("list", nrow(actors))
n_assigned <- 0L

for (row_index in seq_len(nrow(utilities))) {
  option <- utilities[row_index, ]

  if (
    option$id %in% assigned_actors ||
      option$position_id %in% assigned_locations
  ) {
    next
  }

  n_assigned <- n_assigned + 1L
  assignments[[n_assigned]] <- option |>
    select(id, position_id)

  assigned_actors <- c(assigned_actors, option$id)
  assigned_locations <- c(assigned_locations, option$position_id)

  if (n_assigned == nrow(actors)) {
    break
  }
}

actors<-
  actors |> select(id,sex,group,age) |> 
  left_join(bind_rows(assignments), by = "id") |>
  left_join(xy, by = "position_id") |>
  select(-position_id)

```

## The final marriage market
The resulting distribution of actors can be visualized as follows:

```{r}
#| label: assigned-location-plot
#| fig-cap: "Assigned locations. Point size increases with age."
#| fig-align: center
#| fig-width: 6.6
#| fig-height: 6.6
#| out-width: 86%

ggplot(actors, aes(x, y, color = group, shape = sex, size = age)) +
  geom_point(alpha = 0.72) +
  coord_equal(xlim = c(-0.03, 1.03), ylim = c(-0.03, 1.03), expand = FALSE) +
  scale_color_manual(values = group_colors) +
  scale_size_continuous(range = c(1.1, 3.2)) +
  labs(x = "Spatial coordinate x", y = "Spatial coordinate y", color = "Group", shape = "Sex", size = "Age") +
  theme_void(base_size = 12) +
  theme(plot.margin = margin(4, 4, 4, 4))
```

The result is a population with group structure, age structure, and spatial structure. In the next post, I'll uses these actors and locations to create unions based on two-sided partner preferences and a stable matching approach.
