---
title: "Discrete Choice Models for Assortative Mating, Part 1: Simulating Actors and Locations for a Matching Model"
description: "The first note in a small series: create actors, quasi-random locations, group attractors, and a utility-based residential assignment."
date: "2026-07-17"
author: "Benjamin F. Jarvis"
categories:
  - "assortative mating"
  - "simulation"
  - "spatial models"
toc: true
toc-location: right
---

# Intro: Conditional logit models for assortative mating research

There's a been a trend in assortative mating research, particularly in demography and sociology, away from log-linear modeling and towards a more flexible style of discrete choice modeling, primarily based on Daniel McFadden's multinomial logit model.^[Confusingly, this is also referred to as the conditional logistic regression model. This model actually nests what sociologists and political scientists often refer as multinomial logistic regression.]. Examples include the work of Karen Haandrikman^[Haandrikman, K., & van Wissen, L. J. G. (2012). Explaining the Flight of Cupid’s Arrow: A Spatial Random Utility Model of Partner Choice. European Journal of Population, 28(4), 417–439.], Aaron Gullickson^[Gullickson, A. (2021). A counterfactual choice approach to the study of partner selection. Demographic Research, 44, 513–536. https://doi.org/10.4054/DemRes.2021.44.22], and my colleague and student, Jesper Lindmarker^[Lindmarker, J., & Jarvis, B. F. (2025). However Far Away? The Spatial Contingencies of Assortative Mating (4tq8z_v1). SocArXiv. https://doi.org/10.31235/osf.io/4tq8z_v1]. 

That the conditional logit model offers a more general formulation of the log-linear models that have been the "bread and butter" of mobility and assortative mating research has been known, but was not widely acknowledged or leveraged until relatively recently. John A. Logan showed as much in the 1980s^[Logan, J. A. (1983). A Multivariate Model for Mobility Tables. American Journal of Sociology, 89(2), 324–349.], as did Richard Breen in the 1990s^[Breen, R. (1994). Individual Level Models for Mobility Tables and other Cross-Classifications. Sociological Methods & Research, 23(2), 147–173. https://doi.org/10.1177/0049124194023002001].

That this tiny revolution in assortative mating research is taking place outside of economics, where discrete choice models were first developed, is perhaps not too surprising: sociologists are more willing to play fast and loose with statistical methods that have descriptive utility, while not always being so concerned with identification and equilibrium, particularly the stable matching assumptions that often underlie economists work in this area. Using a discrete choice modeling approach largely elides these issues in favor of arriving at parameter estimates that aren't so readibly interpreted in a causal or structural sense, but at least provide rich fodder for sociological interpretation.

So, this research note. My goal is to demonstrate, yet again, the equivalence between conditional logistic regression (i.e., McFadden's choice model) and log-linear analysis, for relatively simple model specifications. But I want to go beyond that and also explore the greater possibilities, as well as the interpretive perils, that go along with this modeling approach. I do so using a simulated dataset representing some intuitive ideas about the factors affecting assortative mating. 

This is just part one. This post sets the scene through simulation; stipulating and then allocating singles to social and spatial positions in a stylized marriage market. In subsequent posts, I'll show how to take these singles and produce matched pairs, how to fit log-linear and conditional logistic regression models to the resulting data, how the parameter estimates compare, and than how these parameter estimates might give us a skewed view of the ``deeper'' parameters that actually drive union formation in the simulated data.

# Building a simulated marriage market

The marriage market we will work with will be characterized by a simplified geometric/geographic space, consisting of quasi-randomly generated positions that are occupied by actors with varying attributes who will then be paired. The pairing step will be covered in another post. This post is just about building the space, generating actors, and assigning actors to positions based on some simplified assumptions about the uneven distribution of different social groups.

# Positions

My goal is to create a semi-regular grid of what might be construed as housing units for actors to occupy. This gives a geographic position to each actor that will influence subsequent matching. A simple way of doing this would be to use independent draws of a set of x and y coordinates from a uniform distribution. However, this creates a set of positions characterized by unusual gaps and overlaps that just doesn't look "right" to my eye. So instead of using pseudo-random numbers, I use quasi-random numbers, in this case Halton sequences, to generate the x and y coordinates. Quasi-random numbers tend to be more evenly spaced that pseudo-random numbers, which feels right for this fictive quasi-geographic set-up.

First things first a few libraries. I prefer `dplyr` for data handling and `ggplot` for plotting. I use `mnorm` for its halton sequence generation:
```{r}
library(dplyr)  
library(ggplot2)  
library(mnorm)
```

We can generate and compare the pseudo- and quasi-random positions as follows:
```{r}
#| label: halton-vs-uniform
#| caption: quasi-geographic positions generated from halton sequences and draws from a uniform distribution

n<-400

xy_sequence <- halton(n = n, base = c(2, 3), start = 1L)

xy<-bind_rows(
tibble(
  position_id = seq_len(n),
  x = xy_sequence[, 1],
  y = xy_sequence[, 2],
  type="halton"
),
tibble(
  position_id = seq_len(n),
  x=runif(n),
  y=runif(n),
  type="uniform"
)
)

ggplot(xy,aes(x=x,y=y)) +
  geom_point() +
  facet_wrap(vars(type),nrow = 1) +
  theme_void() + theme(aspect.ratio=1)

```

As I said, the Halton draws just look a bit better for the present purposes. It resembles a sort of ``social distancing`` heuristic, with positions filling out the space in an irregular, but not overly clumpy way. I save the position generator in a function as follows:

```{r}
make_positions <- function(n) {
  xy_sequence <- halton(n = n, base = c(2, 3), start = 1L)
  tibble(
    position_id = seq_len(n),
    x = position_sequence[, 1],
    y = position_sequence[, 2]
  )
}
```

# Actors

For the actors, I assume they are split into two gendered groups, men and women, with sebsequent unions constituted by opposite gendered pairs. I'll make the market gender balanced for the sake of simplicity, and when all is said and done, each actor will be paired with one and only one opposite gender partner. I stipulate that each partner is characterized by just handful of attributes. First, there will be a group membership, proxying something like race or ethnicity, I'll just use primary colors for the different groups. There will also be age, with the possibility of differences in ages between groups. 


Assortative mating is usually described through attributes of the partners: education, ethnicity, class background, age, religion, or other characteristics. But meetings and choices are also spatial. People do not choose from an abstract population. They choose from people they encounter, and encounters are structured by where people live, study, work, and socialize.

This note builds the first part of a small simulation: actors with group membership, age, sex, attractiveness, and a position in a two-dimensional unit square. The point is not realism. The point is to create a simple data-generating process with enough structure to make later modeling choices visible.

```{r}
#| label: setup
#| message: false
#| warning: false

source("../_matching-simulation.R")

market <- simulate_actors_and_locations(seed = 543)
singles <- market$singles
candidate_positions <- market$candidate_positions
```

## Actors

There are three groups, labeled `a`, `b`, and `c`. The groups differ in size and in mean age. Group `a` is larger and younger; group `c` is smaller and older. The age distributions overlap, however, so group membership is not simply a proxy for age.

```{r}
#| label: actor-summary

singles |>
  count(sex, group)
```

```{r}
#| label: age-summary

singles |>
  group_by(sex, group) |>
  summarize(
    n = n(),
    mean_age = mean(age),
    sd_age = sd(age),
    min_age = min(age),
    max_age = max(age),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

```{r}
#| label: age-plot
#| fig-cap: "Groups differ in mean age, but the simulated age distributions overlap."
#| fig-width: 7
#| fig-height: 4.4

ggplot(singles, aes(age, fill = group)) +
  geom_density(alpha = 0.45, color = NA) +
  facet_wrap(~ sex) +
  scale_fill_manual(values = group_colors) +
  labs(x = "Age", y = "Density", fill = "Group") +
  theme_minimal(base_size = 12)
```

## Candidate Locations

Rather than drawing locations with ordinary pseudo-random numbers, the simulation starts with a quasi-random Halton sequence. This gives a more even coverage of the unit square. The locations are not yet assigned to people. They are just available positions.

```{r}
#| label: candidate-positions
#| fig-cap: "Candidate locations generated from a two-dimensional Halton sequence."
#| fig-width: 5.8
#| fig-height: 5.8

ggplot(candidate_positions, aes(x, y)) +
  geom_point(size = 1.2, alpha = 0.72, color = "#555c60") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  labs(x = "Spatial coordinate x", y = "Spatial coordinate y") +
  theme_void(base_size = 12)
```

## Attractors

The location assignment has two kinds of attractors. Each group has its own group-specific center. Younger actors also feel a pull toward a central youth attractor. The youth pull declines with age.

```{r}
#| label: attractor-plot
#| fig-cap: "Group-specific attractors and the central youth attractor."
#| fig-width: 5.8
#| fig-height: 5.8

ggplot() +
  geom_point(
    data = candidate_positions,
    aes(x, y),
    size = 0.8,
    alpha = 0.28,
    color = "#9aa0a3"
  ) +
  geom_point(
    data = market$group_attractors,
    aes(group_x, group_y, color = group),
    size = 5
  ) +
  geom_point(
    data = market$youth_attractor,
    aes(x, y),
    shape = 8,
    size = 5,
    color = "#1d2224"
  ) +
  geom_text(
    data = market$group_attractors,
    aes(group_x, group_y, label = paste("group", group)),
    nudge_y = 0.055,
    size = 3.4
  ) +
  annotate("text", x = 0.50, y = 0.57, label = "youth", size = 3.4) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  scale_color_manual(values = group_colors) +
  labs(x = "Spatial coordinate x", y = "Spatial coordinate y", color = "Group") +
  theme_void(base_size = 12)
```

For each actor-position pair, the simulation computes a spatial utility. Actors prefer positions closer to their group attractor. Younger actors also prefer positions closer to the youth attractor. Then all person-position pairs are ranked by utility, with a small random tie-breaker, and the simulation walks down the ranked list accepting a pair only when both the person and position are still unassigned.

This is intentionally simpler than a full housing-market model. It creates structured spatial exposure while keeping the assignment mechanism inspectable.

## Assigned Locations

```{r}
#| label: assigned-location-plot
#| fig-cap: "Assigned locations. Point size increases with age."
#| fig-width: 6.3
#| fig-height: 5.8

ggplot(singles, aes(x, y, color = group, shape = sex, size = age)) +
  geom_point(alpha = 0.72) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  scale_color_manual(values = group_colors) +
  scale_size_continuous(range = c(1.1, 3.2)) +
  labs(x = "Spatial coordinate x", y = "Spatial coordinate y", color = "Group", shape = "Sex", size = "Age") +
  theme_void(base_size = 12)
```

The result is a population with group structure, age structure, and spatial structure. The next note uses these actors and locations to create two-sided partner preferences and stable matches.
