---
title: "Does Conditional Logit Represent a Two-Sided Matching Process?"
description: "The fifth note in the series: revisit what a one-sided choice model can and cannot recover from a two-sided stable matching process."
date: "2026-07-17"
author: "Benjamin F. Jarvis"
categories:
  - "conditional logit"
  - "matching"
  - "assortative mating"
toc: true
toc-location: right
---

## The Tension

The simulated matches were generated by a two-sided process. Men ranked women, women ranked men, and the final pairings reflected stable matching under mutual preferences. A conditional-logit model, however, represents the data as if each observed man selected one woman from a choice set.

That representation can be useful. It is also incomplete.

```{r}
#| label: setup
#| message: false
#| warning: false

source("../_matching-simulation.R")
library(survival)

market <- simulate_actors_and_locations(seed = 543)
partner_market <- match_partners(market$men, market$women)
choice_sets <- make_choice_sets(market$men, market$women, partner_market$matched_pairs)
```

## What the Choice Model Sees

The choice model sees the alternatives available to each man and the partner who was ultimately matched with him.

```{r}
#| label: choice-fit

choice_fit <- clogit(
  choice ~ same_group + age_gap + log(distance + 0.01) + factor(candidate_group) + strata(chooser_id),
  data = choice_sets
)

summary(choice_fit)$coefficients
```

This is a valid description of the observed pairing pattern from one side of the market. But it does not explicitly model the fact that some alternatives were unavailable in equilibrium because they preferred someone else.

## A Simple Diagnostic

One way to see the tension is to compare chosen partners with the highest-utility partner from the male side alone. If matching were purely one-sided, these would often coincide. In a two-sided process, they need not.

```{r}
#| label: one-sided-rank

male_rankings <- partner_market$men_to_women |>
  group_by(chooser_id) |>
  arrange(desc(utility), .by_group = TRUE) |>
  mutate(male_side_rank = row_number()) |>
  ungroup()

partner_market$matched_pairs |>
  select(man_id, woman_id) |>
  left_join(
    male_rankings,
    by = c("man_id" = "chooser_id", "woman_id" = "candidate_id")
  ) |>
  summarize(
    mean_male_side_rank = mean(male_side_rank),
    median_male_side_rank = median(male_side_rank),
    share_first_choice = mean(male_side_rank == 1)
  ) |>
  mutate(across(everything(), \(x) round(x, 3)))
```

If many matched partners are not first choices from the male side, that is not a failure of the algorithm. It is the point of two-sided matching: other actors' preferences matter.

## Where This Leaves the Conditional Logit

The conditional-logit model can be a useful descriptive model of assortative matching, especially when the goal is to keep pair-level covariates in view. But it should not be mistaken for the full data-generating process when the underlying process is two-sided.

The practical question is not whether the conditional logit is "true." It is what part of the matching pattern it summarizes, and whether that summary is adequate for the research question.
